---
title: "functions-after.Rmd"
author: "Jai Jot Kaur"
date: "11/10/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

If surveys.csv, species.csv, and plots.csv are not available in your workspace download them, and load them into R using read.csv().

```{r}
surveys <- read.csv("../data-raw/surveys.csv")
species <- read.csv("../data-raw/species.csv")
plots <- read.csv("../data-raw/plots.csv")
```

Combine the surveys and species tables into a single data frame.
```{r}
library(dplyr)
combined <- inner_join(surveys, species, by = "species_id")
```

Write a function that:
Takes three arguments - a data frame (the combined table created before), a genus name, and a species name,
Uses functions from the dplyr package to produce a data frame with two columns: year and count, where count is the number of individuals (i.e., the number of rows) for the species indicated by genus and species in that year,
Returns the resulting data frame. Note that this data frame is a time series by year.

```{r}
time_series <- function(combined, genus_name = "Neotoma", species_name = "albigula") {
 combined %>% 
    select(year, genus, species) %>% 
    filter(genus == genus_name, species == species_name) %>% 
    group_by(year) %>% 
    summarise(count = n()) -> data_frame_result 
  return(data_frame_result)
}
```

```{r}
time_series(combined)
```

Use your function to get the data frame of time series by year for the genus Dipodomys and the species merriami and then make a graph of this time series using ggplot that has year on the x axis, count on the y axis, and displays the data as points.

```{r}
Dipodomys_merriami <- time_series(combined, "Dipodomys", "merriami")
library("ggplot2")
ggplot(data = Dipodomys_merriami, mapping = aes(x = year, y = count)) +  
  geom_point() +
  labs(x = "Year", y = "Count", title = "Time Series by Year for Dipodomys merriami")
```

Use your function to get the data frame time series by year for the genus Chaetodipus and species penicillatus and then make a graph of this time-series using ggplot that has year on the x axis, count on the y axis, and displays the data as blue points (with size = 1) connected by blue lines (with size = 2). Change the x axis label to Year and the y axis label to Number of Individuals.
```{r}
Chaetodipus_penicillatus <- time_series(combined, "Chaetodipus", "penicillatus")
library("ggplot2")
ggplot(data = Chaetodipus_penicillatus, mapping = aes(x = year, y = count)) +  
  geom_point(size = 1, color = "blue") +
  geom_line(size = 2, color = "blue") +
  labs(x = "Year", y = "Number of Individuals", title = "Time Series by Year for Chaetodipus penicillatus")
```


