---
title: "Loops"
author: "Jai Jot Kaur"
date: "11/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
numbers <- c(1, 2, 3, 4, 5)
numbers <- 1:5
numbers <- seq(1, 5, 0.3)
read.csv(file = "../data-raw/species.csv")
mass <- a * length ^ b
for (item in 1:5) {
  print(x = item)
  print("NEW")
}
```

1. The code below prints the numbers 1 through 5 one line at a time. Modify it to print each of these numbers multiplied by 3.

```{r}
numbers <- c(1, 2, 3, 4, 5)
for (number in numbers){
  print(number)
}
```

2. Write a for loop that loops over the following vector and prints out the mass in kilograms (mass_kg = 2.2 * mass_lb)

```{r}
mass_lbs <- c(2.2, 3.5, 9.6, 1.2)
for (items in c(mass_lbs)) {
mass_kg = 2.2 * items
print(items)
}
```


3. Complete the code below so that it prints out the name of each bird one line at a time.

```{r}
birds = c('robin', 'woodpecker', 'blue jay', 'sparrow')
for (i in 1:length(birds)){
  print(birds[i])
}
```


```{r}
our_names <- c("Luna", "Jai", "Azul", "Marcos")
for (items in our_names) {
  print(items)
}
```

4. Complete the code below so that it stores one area for each radius.

```{r}
radius <- c(1.3, 2.1, 3.5)
areas <- vector(mode = "numeric", length = length(radius))
for ( i in 1:length(radius)){
  areas[i] <- pi * radius[i] ^ 2
}
areas
```

5. Complete the code below to calculate an area for each pair of lengths and widths, store the areas in a vector, and after they are all calculated print them out:

```{r}
lengths = c(1.1, 2.2, 1.6)
widths = c(3.5, 2.4, 2.8)
areas <- vector(length = length(lengths))
for (i in 1:length(lengths)) {
  areas[i] <- lengths[i] * widths[i]
}
areas
```

If dinosaur_lengths.csv is not already in your working directory download a copy of the data on dinosaur lengths with species names. Load it into R.

Write a function mass_from_length() that uses the equation mass <- a * length^b to estimate the size of a dinosaur from its length. This function should take two arguments, length and species. For each of the following inputs for species, use the given values of a and b for the calculation from Seebacher 2001:

For Stegosauria: a = 10.95 and b = 2.64
For Theropoda: a = 0.73 and b = 3.63
For Sauropoda: a = 214.44 and b = 1.46
For any other value of species: a = 25.37 and b = 2.49

```{r}
get_mass_from_length <- function(lengths, species) {
  if (species == "Stegosauria") {
    a <- 10.95
    b <- 2.64
  } else if (species == "Theropoda") {
    a <- 0.73
    b <- 3.63
  } else if (species == "Sauropoda") {
    a <- 214.44
    b <- 1.46 
  } else {
    a = 25.37
    b = 2.49
  }
  mass <- a * lengths ^ b
  return(mass)
}
```

Use this function and a for loop to calculate the estimated mass for each dinosaur, store the masses in a vector, and after all of the calculations are complete show the first few items in the vector using head().
Add the results in the vector back to the original data frame. Show the first few rows of the data frame using head().
Calculate the mean mass for each species using dplyr.

```{r}
dino_lengths <- read.csv("../data-raw/dinosaur_lengths.csv")
library(dplyr)
head(dino_lengths)
nrow(dino_lengths)
lengths <- dino_lengths$lengths
species <- dino_lengths$species

all_mass <- vector(length = length(lengths))

all_mass

head(all_mass) 

for (i in 1:length(species)){
  all_mass[i] <- get_mass_from_length(lengths[i], species[i])
}

all_mass

head(all_mass) 

dino_lengths %>% mutate(all_mass = all_mass)

head(all_mass)

dino_lengths %>% 
  group_by(species) %>% 
  summarise(mean_mass = mean(all_mass)) -> species_mass

species_mass
```

