---
title: "Node labels on trees"
author: "Jai Jot Kaur"
date: "11/3/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The data set

Data that comes with an R package.
Working with the `TDbook` package. 

Load the packages and the data package `TDbook`:

```{r}
library(TDbook)
library(ggtree)
library(tidytree)
```

Load the data from teh data package, using the `data()` function.
The data that we ae loading is
- tree_boots
- df_tip_data
- df_inode_data

```{r}
data("tree_boots", "df_inode_data", "df_tip_data")
tree_boots
```

Exercise
1. What is the class of this three objects
```{r}
class(tree_boots)
class(df_inode_data)
class(df_tip_data)
typeof(tree_boots)
typeof(df_inode_data)
typeof(df_tip_data)
```

The class of `tree_boots` is phylogeny.
The class of `df_inode_data` is data frame.
The class of `df_tip_data` is data frame.

2. How many elements does `tree_boots` has
```{r}
str(tree_boots)
```

The number of elements in `tree_boots` is 12.

3. How many rows does `df_tip_data` and `df_inode_data` have.
```{r}
nrow(df_tip_data)
nrow(df_inode_data)
```

The number of rows in `df_tip_data` is 7.
The number of rows in `df_inode_data` is 6.

4. Compare the row numbers of both objects to the lengths of the `$tip.label` and `$node.label` elements of `tree_boots`
```{r}
length(tree_boots$tip.label)
length(tree_boots$node.label)
tree_boots$tip.label %>% length()
tree_boots$node.label %>% length()
```

The length of the `$tip.label` element of `tree_boots` is 7. This is the same as the number of rows in `df_tip_data`. This means that there is a relationship between these two. Each tip label has data
The length of the `$node.label` element of `tree_boots` is 6. This is the same as the number of rows in `df_inode_data`.

5. Join the tree object with the data table that contains tip data and create an object called `tree_joined`.
```{r}
library(tidytree) # not the one from dplyr package
colnames(df_tip_data)
df_tip_data$Newick_label
tree_boots$tip.label
# rename the 'Newick_label" column as "label"
df_mutated <- mutate(df_tip_data, label = Newick_label)
colnames(df_mutated)
df_mutated$Newick_label
df_mutated$label
left_join(tree_boots, df_mutated, by = "label")

df_tip_data

colnames(df_tip_data) -> my_colnames
my_colnames
class(my_colnames)
typeof(colnames)
my_colnames[1]
my_colnames[1] <- "column 1"
my_colnames[10] <- "taxonomy"

colnames(df_tip_data)[10] <- "taxonomy"
df_tip_data

colnames(df_tip_data)[1] <- "label"
df_tip_data

# try to join again tree_boots and df_tip_data

tree_joined <- left_join(tree_boots, df_tip_data)
tree_joined
tree_boots
```

```{r}
tree_joined@phylo
tree_joined@extraInfo$mass_in_kg
```

### Plot node labels into a tree

```{r}
df_tip_data
df_inode_data
```

Join tree_boots to df_inode_data; Hint: Pay attention to the names pf the columns, first change the name of the"newick_label" column to "label":

```{r}
colnames(df_inode_data)[1] <- "label"
colnames(df_inode_data)
tree_data <- left_join(tree_boots, df_inode_data)
tree_data@extraInfo
```

```{r}
ggtree(tree_data) +
  geom_tiplab(fontface = "italic") +
  xlim(0,5) +
  geom_nodelab(mapping = aes(label = vernacularName, color = vernacularName))
```

```{r}
library(dplyr)
species <- read.csv("../data-raw/species.csv")
surveys <- read.csv("../data-raw/surveys.csv")

taxonomy_means <- full_join(surveys, species) %>% 
  filter(!is.na(weight) & !is.na(hindfoot_length)) %>% 
  group_by(species_id) %>% 
  summarise(mean_weight = mean(weight), mean(hindfoot_length))

portal_taxonomy <- read.csv("../data-raw/portal-species-taxonomy.csv")
colnames(portal_taxonomy)

portal_taxonomy_means <- full_join(portal_taxonomy, taxonomy_means) 

library(tidytree)
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
colnames(portal_taxonomy_means)
portal_joined <- tidytree:::left_join.phylo(portal_tree, portal_taxonomy_means)
```

Homework
Add node labels to your two tree plots of Portal tree data with average weight and hindfoot length. Use the column "taxa" both as label and fill color.
1. Join portal_tree and the dataframe with average weights and hindfoot lengths, and taxa data.
2. Plot it
3. Add tip labels colored by taxa
4. Add node labels

```{r}
ggtree(portal_joined) +
  geom_tiplab(mapping = aes(color = taxa), fontface = "italic", size = 2) +
  xlim(0,15) +
  geom_nodelab(mapping = aes(color = taxa), size =  2)
```

