---
title: "portal phylogenies.Rmd"
author: "Jai Jot Kaur"
date: "10/25/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Add an intro to the newick format
## A newick file is a text file that demonstrates a representation for a phylogenetic tree. This text file allows the easy accessibility and reading of the tree by the computer. An inner node is represented by a pair of paranthesis, a tip is represented by names or character strinngs, a lineage is represented by a comma which also allwos separation of the nodes and tups, and, finally, the semicolon is representative of the end of the tre; therefore, multiple semicolons in a newick file may represent multiple trees. This is useful as image files for the data are not able to be read by a cimputer, so a text file, such as the nwick file, is necessary for the computer to easily read the file

# The package 'ape'

```{r}
# install.packages("ape")
library(ape)
```

To read a newick tree from a local file:

```{r}
portal_tree <- read.tree("../data-raw/portal-tree.tre")
portal_tree
```

Read a newick from a URL 

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```

# The structure of a tree in R

```{r}
surveys <- read.csv("../data-raw/surveys.csv")
class(portal_tree)
length(portal_tree)
length(surveys)
colnames(surveys)
colnames(portal_tree)
names(portal_tree)
portal_tree$Nnode
portal_tree["tip.label"]
portal_tree[[3]]
summary(portal_tree)
str(portal_tree)
class(portal_tree$edge)
portal_tree$edge
```

```{r}
str(surveys)
```

```{r}
plot.phylo(x = portal_tree)
```

An extension of ggplot for phylogenetic visualization: `ggtree`
To install packages from the Bioconductor repo, we need a CRAN package called BiocManager

```{r}
# install.packages("Biocmanager")
library(BiocManager)
# install("ggtree")
library(ggtree)
```

```{r}
ggtree(portal_tree)
```

This is equivalent to:

```{r}
ggplot(portal_tree, aes(x, y)) +
  geom_tree() +
  theme_tree()
```

Plot the small tree of five species of primates and include a scale what is the difference between the portal and the small tree 
```{r}
ggtree(small_tree) +
  geom_treescale()
```

```{r}
class(small_tree)
length(small_tree) # gives the number of 
names(small_tree)
small_tree$Nnode
small_tree["tip.label"]
head(small_tree[[3]])
summary(small_tree)
str(small_tree) # $ to access elements
head(small_tree$edge)
```

```{r}
ggtree(portal_tree) +
geom_treescale()
ggtree(small_tree) +
geom_treescale()
```


```{r}
names(small_tree)
```

- a difference in number of tips (43 vs 5)
- an edge length in 'small_tree', but it has no node labels

```{r}
small_tree$edge.length
small_tree$node.label
portal_tree$node.label
```

## Add tip labels and node labels

Because a plot is nothing with no labels:

```{r}
ggtree(portal_tree) + 
  geom_treescale() +
  geom_tiplab(size = 1, color = "purple", fontface = "italic")
```

Add a limit to the plot so we can see the labels fully: 

```{r}
ggtree(portal_tree) + 
  geom_treescale() +
  geom_tiplab(size = 1, color = "purple", fontface = "italic") +
  xlim(NA,200)
```

Do this for small_tree for homework

Get branch lengths

```{r}
branching.times(small_tree)
ggtree(small_tree) + 
  geom_treescale() +
  geom_tiplab(size = 4, color = "purple", fontface = "italic") +
  xlim(NA,1.15)
```

```{r}
ggtree(portal_tree, layout = "fan", open.angle = 90) +
  geom_tiplab(fontface = "italic")
```

```{r}
# install.packages("aplot")

library(aplot)

ggtree(portal_tree, layout = "fan", open.angle = 45) -> plot1

plot_list(plot1, plot1, labels = c("Circular Tree of Portal Species", "The same"), tag_size = 10)
```

Part 1:
```{r}
# install.packages("igraph")
# library(igraph)
p1 <- ggtree(portal_tree, layout="rectangular")
p2 <- ggtree(portal_tree, layout="ape")
p3 <- ggtree(portal_tree, layout="roundrect")
p4 <- ggtree(portal_tree, layout="slanted")
p5 <- ggtree(portal_tree, layout="ellipse")
p6 <- ggtree(portal_tree, layout="dendrogram") # pay attention to spelling
p7 <- ggtree(portal_tree, layout="circular")
p8 <- ggtree(portal_tree, layout="radial")
p9 <- ggtree(portal_tree, layout="fan", open.angle = 90)
p10 <- ggtree(portal_tree, layout="equal_angle")
p11 <- ggtree(portal_tree, layout="daylight")
p12 <- ggtree(portal_tree, layout="unrooted")
```

Part 2:
Create a subplot for each tree.
```{r}
plot_list(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12, tag_levels = "a")
```

Part 3: Is the tree representation rooted or unrooted

```{r}
# create a vector of labels for my 12 plots:
my_labels <- paste(letters[1:12], "=", "rooted")
my_labels
plot_list(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,
          labels = my_labels, tag_size = 5)
```

# Connecting a phylogeny with data from a table

```{r}
taxonomy <- read.csv("../data-raw/portal-species-taxonomy.csv")
surveys <- read.csv("../data-raw/surveys.csv")
species <- read.csv("../data-raw/species.csv")
```

```{r}
intersect(colnames(species), colnames(surveys))
library(dplyr)
inner_join(species, surveys, by = "species_id")
```

```{r}
left_join(portal_tree, taxonomy, by = "label")
head(taxonomy)
taxonomy$label
```

```{r}
portal_tree$tip.label
```

Replace the tip labels from our original portal tree with a new vector of tip labels:
```{r}
tip_labels <- portal_tree$tip.label
tip_labels
tip_labels_cleaned <- gsub("_", " ", tip_labels)
portal_tree$tip.label
```

```{r}
tree_data <- left_join(portal_tree, taxonomy, by = "label")
tree_data
class(tree_data)
```

Color by a variable in our column

```{r}
ggtree(tr = tree_data, mapping = aes(color = taxa)) +
  geom_tiplab(size = 2.5, fontface = "italic") +
  xlim(0,100)
ggsave(filename = "../fig/portal-tree-1.png")
```
### Troubleshooting tree and dat ajoins

```{r}
portal_tree$tip.label
taxonomy$label
match(taxonomy$label, portal_tree$tip.label)
match(portal_tree$tip.label, taxonomy$label)
```

```{r}
portal_tree <- read.tree(file = "../data-raw/portal-tree.tre")
portal_tree
class(portal_tree)
typeof(portal_tree)
str(portal_tree)
```
```{r}
tax  <- read.csv(file = "../data-raw/portal-species-taxonomy.csv")
write.tree(phy = portal_tree, file = "../data-raw/portal-tree.tre")
write.csv(taxonomy, file = "../data-raw/taxonomy-test.csv", row.names = FALSE)
```

```{r}
library(tidytree)
left_join(portal_tree, taxonomy, by = "label")
## the join failed!!
# we need to check our species labels on the tree and on the data table:
head(taxonomy)
taxonomy$label
```

```{r}
tip_label <- small_tree$tip.label
left_join(small_tree, small_tree, by = "tip.label")
```

### Exercise - homework

Exercise: A taxonomy table for small_tree
Find the appropriate scientific group labels for each genus in small_tree using this tree as guide.
Create a data frame with 3 columns:
a "label" column with the names of the tip labels of small_tree. Tip: extract the element "tip.label" from your phylo object to get a vector of tip labels that you can then join to the other vectors to create a data frame.
a "taxa" column with the scientific names of the group that each genus belongs to.
a "common_name" column with the common names of the group that each genus belongs to. Tip: use the function c() to create the vectors that will be columns "taxa" and "common_name"
Join your tree and your table using left_join().
Create two different tree plots using taxa and common name to color the tips of the tree.

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
branching.times(small_tree)
ggtree(small_tree) + 
  xlim(0,1.5) +
  geom_tiplab(fontface = "italic")

label <- small_tree$tip.label
taxa <- c("Hominini", "Ponginae", "Cercopithecoidea", "Platyrrhini", "Lorisoidea")
common_name <- c("Human", "Orangutan", "Macaques", "Spider Monkey", "Lesser Bushbaby")

data.frame(label, taxa, common_name)
```

```{r}
# install.packages("ggimage")
library(ggimage)
library(ggtree)
# install.packages("TDbook")
library(TDbook)
library(tidytree)
data("tree_boots")
data("df_tip_data")
data("df_inode_data")
trr_boots
# column "Newick_label" change to "label  in df_tip_data and in df_inode_data"
```

Using mutate
```{r}
renamed_tip_data <- df_tip_data %>% 
  mutate(label = Newick_label)
colnames(renamed_tip_data)
renamed_inode_data <- df_inode_data %>% 
  mutate(label = newick_label)
colnames(renamed_inode_data)
renamed_inode_data$label
```

Using colnames function

```{r}
colnames(df_tip_data)[1] <- "label"
colnames(df_tip_data)
df_inode_data
colnames(df_inode_data)[1] <- "label"
colnames(df_inode_data)
```

Join `tree_boots` to `df_tip_data`

```{r}
names(tree_boots)
colnames(df_tip_data)
intersect(names(tree_boots), colnames(df_tip_data))
tree_joined <- left_join(tree_boots, df_tip_data, by = "label")
tree_joined
```

Plot some data to the tips of `tree_boots`

```{r}
colnames(df_tip_data)
df_tip_data$rank

ggtree(tree_joined) + 
  geom_tippoint(mapping = aes(shape = trophic_habit,
                              color = trophic_habit,
                              size = mass_in_kg)) +
  geom_tiplab(fontface = "italic") +
  xlim(0, 5)

df_tip_data
```

Exercise: Mapping weight data from surveys CSV table to the portal tree
1. Get the average weight and hindfoot length per species.
2. Create a new data frame that contains the taxonomy data plus the averaged data per species that you got on last question.
3. Create two plots with data on the tips, one with the average weight and the other with average hindfoot length. Make sure to also add tip labels, formatted in italics.

```{r}
library(dplyr)
colnames(surveys)
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id) %>% 
  summarise(mean_weight = mean(weight)) -> species_weight

nrow(species_weight)
spp_weight <- inner_join(species, species_weight)
spp_weight

colnames(taxonomy)
taxonomy_weight <- full_join(taxonomy, spp_weight)
taxonomy_weight

tree_join <- left_join(portal_tree, taxonomy_weight, by = "label")

tree_join

ggtree(tree_join) +
  geom_tiplab(fontface = "italic", size = 2) + 
  xlim(0,20) +
  geom_tippoint(mapping = aes(shape = "circle",
                              size = mean_weight,
                              color = taxa))
```

```{r}
library(dplyr)
colnames(surveys)
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id) %>% 
  summarise(mean_hindfoot_length = mean(hindfoot_length)) -> species_hindfoot_length

nrow(species_hindfoot_length)
spp_hindfoot_length <- inner_join(species, species_hindfoot_length)
spp_hindfoot_length

colnames(taxonomy)
taxonomy_hindfoot_length <- full_join(taxonomy, spp_hindfoot_length)
taxonomy_hindfoot_length

tree_join <- left_join(portal_tree, taxonomy_hindfoot_length, by = "label")

tree_join

ggtree(tree_join) +
  geom_tiplab(fontface = "italic", size = 2) + 
  xlim(0,20) +
  geom_tippoint(mapping = aes(shape = "circle",
                              size = mean_hindfoot_length,
                              color = taxa))
```



