---
title: "Review - uhuru"
author: "Jai Jot Kaur"
date: "12/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A function for the UHURU data set

Explain what each line of code in the body of the function is doing. Add the explanations to your Rmd file as comments, before each line of code.
```{r}
# this creates a function named report_rsquared which sonsists of the arguments data, species, and formula
report_rsquared <- function(data, species, formula){
  # this creates a subset named subset which filters the data to only include the species ANT
  subset <- dplyr::filter(data, ANT == species)
  # this creates a vector knoene as test which allows the linear models to be fit to data frames
  test <- lm(formula, data = subset)
  # this rounds the value for r squared from the summary of trst by 3 decimal places to acquire a value
  rsquared <- round(summary(test)$r.squared, 3)
  # this function creates a data frame which specifies the values for species and the r squared value
  output <- data.frame(species = species, r2 = rsquared)
  # this function returns the object for the output 
  return(output)
}
```

Execute the function using the UHURU data and specifying species = "CM" and formula = "AXIS1~CIRC".
```{r}
uhuru_data <- read.csv(file = "https://esapubs.org/archive/ecol/E095/064/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
head(uhuru_data)
report_rsquared(data = uhuru_data, species = "CM", formula = "AXIS1~CIRC")
```

Modify the function so that it also determines if() the rsquared is significant based on a given threshold. The modified function should return() the species, rsquared and a significance value of "S" for a relationship with an rsquared > threshold or "NS" for an rsquared < threshold.

```{r}
report_rsquared <- function(data, species, formula, threshold){
  subset <- dplyr::filter(data, ANT == species)
  test <- lm(formula, data = subset)
  rsquared <- round(summary(test)$r.squared, 3)
  
  if (rsquared > threshold) {
    print("Significant")
    sign <- "S"
  } else if (rsquared < threshold) {
    print("Non significant")
    sign <- "NS"
  }
  output <- data.frame(species = species, r2 = rsquared, significance = sign)
  return(output)
}
```

Execute your modified function for species of "CM", "CS", and "TP" given a threshold = 0.667.
```{r}
report_rsquared(data = uhuru_data, species = "CM", formula = "AXIS1~CIRC", threshold = 0.667)
# "Significant"; r2 = 0.866

report_rsquared(data = uhuru_data, species = "CS", formula = "AXIS1~CIRC", threshold = 0.667)
# "Non significant"; 0.437

report_rsquared(data = uhuru_data, species = "TP", formula = "AXIS1~CIRC", threshold = 0.667)
# "Significant"; 0.701
```




