---
title: "Review"
author: "Jai Jot Kaur"
date: "11/29/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load or Download a File?

Write a conditional statement that checks if surveys.csv exists in the working directory, if it doesn’t then downloads it from https://ndownloader.figshare.com/files/2292172 using download.file(), and finally loads the file into a data frame and displays the first few rows using the head() function. The url needs to be in quotes since it is character data.

```{r}
survey.list <- list.files("/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw")
any(survey.list)
library(stringr) 
str_detect(list.files("/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw"), "surveys.csv")
is.element("surveys.csv", list.files("/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw"))
??str_detect
getwd()
"a" == c("a", "b", "c")
?list.files

if (is.element("surveys.csv", survey.list)) {
  print("The file surveys.csv is already downloaded")
  } else {
    print("The file is not downloaded")
  }
# downloads it from https://ndownloader.figshare.com/files/2292172 using download.file()
    download.file("https://ndownloader.figshare.com/files/2292172", "../data-raw/surveys-download.csv")
# loads the file into a data frame
    surveydataframe <- read.csv("../data-raw/surveys-download.csv")
# displays the first few rows using the head() function 
     head(surveydataframe)
```

Make a version of this conditional statement that is a function, where the name of the file is the first argument and the link for downloading the file is the second argument.
The function should return the resulting data frame.
Add some documentation to the top of the function describing what it does.
Call this function using “species.csv” as the file name and https://ndownloader.figshare.com/files/3299483 as the link. Print the first few rows of the resulting data frame using head().

```{r}
# This function tests if a file is in the data raw directory and if not, it downloads it and reads it as data frame

reading_csv <- function(file_name, link) {
  
# 1. Test if file is in the data-raw folder 
  # file_name <- "species.csv"
  test <- !is.element(file_name, list.files(path = "../data-raw"))
  download.file(link, "../data-raw/file")
  
# 2. If test is FALSE, download the file
  if(test) {
    
    # Option 1: save it with a random name:
    # download.file(url = file_link, destfile = "../data-raw/temporary.csv")
    # result <-  read.csv(file = "../data-raw/temporary.csv")
    
    # Option 2: save it eith the name given in file name:
    destination_file <- stringr:: str_c("../data-raw", file_name)
    download.file(url = link, destfile = destination_file)
    result <- read.csv(file = destination_file)
   return(result)
  }
  
}

```

```{r}
reading_csv(file_name = "species.csv", link = "https://ndownloader.figshare.com/files/3299483")
```

## Multi-file Analysis

If individual_collar_data.zip is not already in your working directory download the zip file using download.file()
```{r}
list_of_files <- list.files(path = "../data-raw/.")
files_present <- is.element("individual_collar_data.zip", list_of_files)

file_name <- "individual_collar_data.zip"


if(!files_present) {
  download.file("http://www.datacarpentry.org/semester-biology/data/individual_collar_data.zip", "../data-raw/individual_collar_data.zip")
}

library(stringr)
str_c("../data-raw/", file_name)
```

Unzip it using unzip()
```{r}
# dir.create(path = "../data-raw/individual_collar_data")
unzip("../data-raw/individual_collar_data.zip", exdir = "../data-raw/individual_collar_data")
```

Obtain a list of all of the files with file names matching the pattern "collar-data-.*.txt" (using list.files())
```{r}
collar_data_files <- list.files("/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw", "collar-data-.*.txt")
```

Use a loop to load each of these files into R and make a line plot (using geom_path()) for each file with long on the x axis and lat on the y axis. Graphs, like other types of output, won’t display inside a loop unless you explicitly display them, so you need put your ggplot() command inside a print() statement. Include the name of the file in the graph as the graph title using labs()
```{r}
# load file
read.csv(file = "../data-raw/individual_collar_data/collar-data-J10-2016-02-26.txt")

library(ggplot2)

for (i in collar_data_files) {
  print(i)
  collar_data_table <- read.csv(file = i)
  # print(head(collar_data_table))
  ggplot(collar_data_table, aes(x = long, y = lat)) +
    labs(title = i) +
    geom_path() -> collar_data_graph
  print(collar_data_graph)
}
```

Add code to the loop to calculate the minimum and maximum latitude in the file, and store these values, along with the name of the file, in a data frame. Show the data frame as output.

Solution 1, using empty vectors
```{r}
all_min <- vector()
all_max <- vector()
length(all_min)
length(all_max)
file_name <- vector()
?read.csv
for (i in collar_data_files) {
  print(i)
  collar_data_table <- read.csv(str_c("../data-raw/individual_collar_data/", i))
  file_name <- c(file_name, i)
  min_lat <- min(collar_data_table[, "lat"])
  # min_lat <- min(collar_data_table$lat)
  all_min <- c(all_min, min_lat)
  print(all_min)
  max_lat <- max(collar_data_table[, "lat"])
  # max_lat <- max(collar_data_table$lat)
  all_max <- c(all_max, max_lat)
  print(all_max)
}

minmax_dataframe <- data.frame(file_name, all_min, all_max)
minmax_dataframe
```

Solution 2, using vectors of a predetermined length

```{r}

#  create vectors of a certain length 

all_min_lat <- vector(mode = "numeric", length = length(collar_data_files))
all_max_lat <- all_min_lat
all_file_name <- all_min_lat
length(all_min)
length(all_max)
all_file_name
for (i in 1:length(collar_data_files)) {
  file_name_and_path <- str_c("../data-raw/individual_collar_data/", collar_data_files[i])
  all_file_name[i] <- file_name_and_path
  print(file_name_and_path)
  collar_data_table <- read.csv(file = file_name_and_path)
  min_lat <- min(collar_data_table$lat)
  all_min_lat[i] <- min_lat
  max_lat <- max(collar_data_table$lat)
  all_max_lat[i] <- max_lat
  
  print(all_max_lat)
  print(all_min_lat)
}
length(file_name_and_path)
vector_minmaxdataframe <- data.frame(all_file_name, all_min_lat, all_max_lat)
vector_minmaxdataframe
```

Extra. If you’re interested in seeing another application of for loops, check out the code used to simulate the data for this exercise using for loops.
```{r}
individuals = paste(c('A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J'), c(1:10), sep = "")
for (individual in individuals) {
    lat = vector("numeric", 24)
    long = vector("numeric", 24)
    lat[1] = rnorm(1, mean = 26, sd = 2)
    long[1] = rnorm(1, mean = -35, sd = 3)
    for (i in 2:24) {
        lat[i] = lat[i - 1] + rnorm(1, mean = 0, sd = 1)
        long[i] = long[i - 1] + rnorm(1, mean = 0, sd = 1)
    }
    times = seq(from=as.POSIXct("2016-02-26 00:00", tz="UTC"),
                to=as.POSIXct("2016-02-26 23:00", tz="UTC"),
                by="hour")  
    df = data.frame(date = "2016-02-26",
                    collar = individual,
                    time = times,
                    lat = lat,
                    long = long)
    write.csv(df, paste("collar-data-", individual, "-2016-02-26.txt", sep = ""))
}
zip("data/individual_collar_data.zip", list.files(pattern = "collar-data-[A-Z][0-9]+-.*"))
```

