---
title: "Review Energy Conversion Challenge"
author: "Jai Jot Kaur"
date: "12/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Write a function called convert_energy_units(energy_value, input_unit, output_unit) to convert units between the following energy values:
Joules(J),
Kilojoules(KJ),
Calories(CAL),
and Kilocalories (KCAL; this unit is used for labeling the amount of energy contained in food). A Kilojoule is 1000 Joules, a Calorie is 4.1868 Joules, a Kilocalorie is 4186.8 Joules. An example of a call to this function would look like:

```{r}
convert_energy_units <- function(energy_value, input_unit, output_unit) {
  if (input_unit == "J" & output_unit == "CAL") {
    energy_in_cal <- energy_value/4.184
    print(energy_in_cal)
  } else if (input_unit == "J" & output_unit == "KJ") {
  energy_in_kj <- energy_value/1000
  print(energy_in_kj)
  } else if (input_unit == "J" & output_unit == "KCAL") {
  energy_in_kcal <- energy_value/4184
  print(energy_in_kcal)
  } else if (input_unit == "CAL" & output_unit == "J") {
   energy_in_j <- energy_value * 4.1868  
   print(energy_in_j)
  } else if (input_unit == "KCAL" & output_unit == "J") {
    energy_in_j <- energy_value * 4186.8
    print(energy_in_j)
  }  else if (input_unit == "KJ" & output_unit == "J") { 
    energy_in_j <- energy_value*1000
    print(energy_in_j)
  } else {
    print("Sorry, I don’t know how to convert")
  }
}

energy_in_cal <- 200
energy_in_j <- convert_energy_units(energy_in_cal, "CAL", "J")
# 837.36
```

Use your function to answer the following questions:
What is the daily metabolic energy used by a human (~2500 KCALs) in Joules.
```{r}
metabolic_energy_in_kcal <- 2500
metabolic_energy_in_j <- convert_energy_units(metabolic_energy_in_kcal, "KCAL", "J")
# 10467000
```

How many times more energy does a common seal use than a human? The common seal uses ~52,500 KJ/day (Nagy et al. 1999). Use the daily human metabolic cost given above.
```{r}
seal_energy_in_kj <- 52500
seal_energy_in_j <- convert_energy_units(seal_energy_in_kj, "KJ", "J")
# 52500000
seal_energy_in_j/metabolic_energy_in_j
# 5.015764
```

How many ergs (ERG) are there in one kilocalorie. Since we didn’t include the erg conversion this should trigger our ‘don’t know how to convert’ message.
```{r}
one_kcal <- 1
energy_in_erg <- convert_energy_units(one_kcal, "KCAL", "ERG")
# "Sorry, I don’t know how to convert"
```

Make the function more efficient, and instead of writing an individual conversion between each of the different units (which requires 12 if statements) you could choose to convert all of the input units to a common scale and then convert from that common scale to the output units. This approach is especially useful if you need to add new units later.

Solution1: Creating two functions independently and then combining them
```{r}
# difference between print and return inside function
# cleaner way to print message inside function is the function message. print function can be sued to generate an output for the function; tis is not efficent as using a simple return. Using the return can allow more complcicated things to be returned such as lists. 
convert_to_j <- function(energy_value, input_unit, output_unit) {
  
  # Convert Joules to output_unit
  if (input_unit == "CAL") {
   energy_in_j <- energy_value * 4.1868  
   return(energy_in_j)
  } else if (input_unit == "KCAL") {
    energy_in_j <- energy_value * 4186.8
    return(energy_in_j)
  }  else if (input_unit == "KJ") { 
    energy_in_j <- energy_value*1000
    return(energy_in_j)
  }  else {
    message("Sorry, I don’t know how to convert", input_unit)
    return(NA)
  }
}
  
convert_energy_efficient <- function(energy_value, input_unit, output_unit) {
  
   # Convert input to J
 energy_in_joules <- convert_to_j(energy_value, input_unit) 
# convert Joules to output_unit
   if (output_unit == "CAL") {
    energy <- energy_in_joules/4.1868 
  } else if (output_unit == "KJ") {
  energy <- energy_in_joules/1000
  } else if (output_unit == "KCAL") {
  energy <- energy_in_joules/4186.8
  }
 return(energy)
}

convert_energy_efficient(1200, "CAL", output_unit = "KCAL")
# 1.2
```

Solution22: Writing all the code inside one single function
