---
title: "Review Energy Conversion Challenge"
author: "Jai Jot Kaur"
date: "12/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Write a function called convert_energy_units(energy_value, input_unit, output_unit) to convert units between the following energy values:
Joules(J),
Kilojoules(KJ),
Calories(CAL),
and Kilocalories (KCAL; this unit is used for labeling the amount of energy contained in food). A Kilojoule is 1000 Joules, a Calorie is 4.1868 Joules, a Kilocalorie is 4186.8 Joules. An example of a call to this function would look like:

```{r}
convert_energy_units <- function(energy_value, input_unit, output_unit) {
  if (input_unit == "J" & output_unit == "CAL") {
    energy_in_cal <- energy_value/4.184
    print(energy_in_cal)
  } else if (input_unit == "J" & output_unit == "KJ") {
  energy_in_kj <- energy_value/1000
  print(energy_in_kj)
  } else if (input_unit == "J" & output_unit == "KCAL") {
  KCAL <- energy_value/4184
  print(energy_in_kcal)
  } else if (input_unit == "CAL" & output_unit == "J") {
   energy_in_j <- energy_value * 4.1868  
   print(energy_in_j)
  } else if (input_unit == "KCAL" & output_unit == "J") {
    energy_in_j <- energy_value * 4186.8
    print(energy_in_j)
  }  else if (input_unit == "KJ" & output_unit == "J") { 
    energy_in_j <- energy_value*1000
    print(energy_in_j)
  } else {
    print("Sorry, I don’t know how to convert")
  }
}

energy_in_cal <- 200
energy_in_j <- convert_energy_units(energy_in_cal, "CAL", "J")
# 837.36
```

Use your function to answer the following questions:
What is the daily metabolic energy used by a human (~2500 KCALs) in Joules.
```{r}
metabolic_energy_in_kcal <- 2500
metabolic_energy_in_j <- convert_energy_units(metabolic_energy_in_kcal, "KCAL", "J")
# 10467000
```

How many times more energy does a common seal use than a human? The common seal uses ~52,500 KJ/day (Nagy et al. 1999). Use the daily human metabolic cost given above.
```{r}
seal_energy_in_kj <- 52500
seal_energy_in_j <- convert_energy_units(seal_energy_in_kj, "KJ", "J")
# 52500000
seal_energy_in_j/metabolic_energy_in_j
# 5.015764
```

How many ergs (ERG) are there in one kilocalorie. Since we didn’t include the erg conversion this should trigger our ‘don’t know how to convert’ message.
```{r}
one_kcal <- 1
energy_in_erg <- convert_energy_units(one_kcal, "KCAL", "ERG")
# "Sorry, I don’t know how to convert"
```

