---
title: "small mammals dataset"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Portal Project Teaching Dataset

It is an ecological experiment made in the Chihuahuan deser in Arizona near the city of Portal

Later I will add a picture of the Chihuahuan desert
https://www.google.com/url?sa=i&url=https%3A%2F%2Fwww.desertusa.com%2Fchihuahuan-desert.html&psig=AOvVaw1ccqb81iiGHfvmvPe9xZXL&ust=1665611083984000&source=images&cd=vfe&ved=0CAwQjRxqFwoTCPj2uP2S2foCFQAAAAAdAAAAABAD

For all the data manipulations we are going to be using the "dplyr" package.

To install packages we use the function 'install_packages()': But to use the functions inside a package, we need to load them to our environment, we do this with the function 'library()'.

```{r}
library(dplyr)
```

# Read the data table

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
colnames(surveys)
# View(surveys)
plots <- read.csv(file = "../data-raw/plots.csv")
head(plots)
# View(plots)
species <- read.csv(file = "../data-raw/species.csv")
head(species)
# View(species)
```

# Subset a data table

Using the dplyr function 'subset()' :

```{r}
# select(surveys, year, species_id, weight)
# surveys_short <- select(surveys, year, species_id, weight)
# head(surveys_short)
```

Other ways to subset a data table:

```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,"year"]
```

## Add columns to a data table

We can use the function 'mutate()' 

Let's say that we want to add on extra column that has weight divided by 10

Using the classic approach

```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10

colnames(surveys)
```
```{r}
# surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```

## Rearrange/reorder a data table by some value

We will use the function 'arrange()'

```{r}
# surveys_arranged <- arrange(surveys, weight)

# tail(surveys_arranged, 100)
```

To arrange in descending order

```{r}
# arrange(surveys, desc(weight))
```

## Filter data 

To filter unwanted data or by some specific value in our data we use the function 'filter()'

```{r}
# filter(surveys, species_id == "DS")
# filter(surveys, species_id != "DS")
```

We can also filter by multiple conditions:

```{r}
# temp_table <- filter(surveys, species_id == "DS", year >= "1998")
# tail(temp_table, 100)
```

The logical operator '&' and '|' mean "AND" and "OR"

```{r}
# filter(surveys, species_id == "DS" & year >= "1998")
```

```{r}
# filter(surveys, species_id == "DS" | year >= "1998")
```

Let's filter for values that are NA, we need the function 'is.na'

```{r}
# is.na(surveys)
# surveys$weight
# filter(surveys, !is.na(weight))
```

# Load surveys.csv into R using read.csv().

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
```

# Use select() to create a new data frame object with just the year, month, day, and species_id columns in that order.

```{r}
surveys_new <- select(surveys, year, month, day, species_id)
```

# Use mutate(), select(), and filter() with !is.na() to create a new data frame with the year, species_id, and weight in kilograms of each individual, with no null weights. The weight in the table is given in grams so you will need to create a new column for weight in kilograms by dividing the weight column by 1000.

```{r}
surveys_subset <- filter(surveys, !is.na(weight))
surveys_subset <- mutate(surveys_subset, weight_kg = weight/1000)
surveys_subset <- select(surveys_subset, year, species_id, weight_kg)
filter(surveys_subset, species_id == "SH")
```
# Pipes

`%>%` - `magrittr`  package required
'|>'  - inbuilt in R, but very new,s o you need an updated version of RStudio

```{r}
mean(x = c(1, 3, 65, 38, 100)) ## sequential code
c(1, 3, 65, 38, 100) %>% mean()

sqrt(mean (x = c(1, 3, 65, 38, 100)))

c(1, 3, 65, 38, 100) %>% mean () %>% sqrt()
```
## Use mutate(), select(), and filter() with is.na() to create a new data frame with the year, species_id, and weight in kilograms of each individual, with no null weights.

```{r}
surveys1 <- surveys %>% 
  mutate(weight_kg = weight/1000) %>%
  select(year, species_id, weight_kg) %>%
  filter(!is.na(weight_kg))
colnames(surveys1)
is.na(surveys1$weight_kg) %>%
  any()
```

## Use the filter() and select() to get the year, month, day, and species_id columns for all of the rows in the data frame where species_id is SH.

```{r}
surveys2 <- surveys %>% 
  select(year, month, day, species_id, weight) %>%
  filter(species_id == "SH") %>%
  filter(!is.na(weight)) %>%
  select(year, month, day, species_id)

surveys2$species_id == "SH"
surveys[surveys2$specis_id == "SH",]
colnames(surveys2)
nrow(surveys2)
```

## Obtain the data for only DS, sorted by year, with only the year and and weight columns

```{r}
ds_weight_by_year <- surveys %>%
  filter(species_id == "DS") %>%
  filter(!is.na(weight)) %>%
  arrange(year) %>%
  select(year, weight)
nrow(ds_weight_by_year)
```

## Data Aggregation

```{r}
surveys_by_year <- group_by(surveys, year)

surveys_by_year
```
```{r}
summarize(surveys_by_year, abundance = n())
```


## Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID.

```{r}
surveys_by_species_id <- surveys %>%
  group_by(species_id) %>%
  summarize(abundance = n())
surveys_by_species_id
```

## Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID in each year.
```{r}
surveys_by_species_id_and_year <- surveys %>%
  group_by(species_id, year) %>%
  summarize(abundance = n())
surveys_by_species_id_and_year
```

## Use the filter(), group_by(), and summarize() functions to get the mean mass of species DO in each year.
```{r}
do_mean_mass_by_year <- surveys %>%
  filter(species_id == "DO") %>%
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  summarize(mean_weight_by_year = mean(weight)) 
do_mean_mass_by_year
```


