---
title: "small mammals dataset"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Portal Project Teaching Dataset

It is an ecological experiment made in the Chihuahuan deser in Arizona near the city of Portal

Later I will add a picture of the Chihuahuan desert
https://www.google.com/url?sa=i&url=https%3A%2F%2Fwww.desertusa.com%2Fchihuahuan-desert.html&psig=AOvVaw1ccqb81iiGHfvmvPe9xZXL&ust=1665611083984000&source=images&cd=vfe&ved=0CAwQjRxqFwoTCPj2uP2S2foCFQAAAAAdAAAAABAD

For all the data manipulations we are going to be using the "dplyr" package.

To install packages we use the function 'install_packages()': But to use the functions inside a package, we need to load them to our environment, we do this with the function 'library()'.

```{r}
library(dplyr)
```

# Read the data table

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
colnames(surveys)
# View(surveys)
plots <- read.csv(file = "../data-raw/plots.csv")
head(plots)
# View(plots)
species <- read.csv(file = "../data-raw/species.csv")
head(species)
# View(species)
```

# Subset a data table

Using the dplyr function 'subset()' :

```{r}
# select(surveys, year, species_id, weight)
# surveys_short <- select(surveys, year, species_id, weight)
# head(surveys_short)
```

Other ways to subset a data table:

```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,"year"]
```

## Add columns to a data table

We can use the function 'mutate()' 

Let's say that we want to add on extra column that has weight divided by 10

Using the classic approach

```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10

colnames(surveys)
```
```{r}
# surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```

## Rearrange/reorder a data table by some value

We will use the function 'arrange()'

```{r}
# surveys_arranged <- arrange(surveys, weight)

# tail(surveys_arranged, 100)
```

To arrange in descending order

```{r}
# arrange(surveys, desc(weight))
```

## Filter data 

To filter unwanted data or by some specific value in our data we use the function 'filter()'

```{r}
# filter(surveys, species_id == "DS")
# filter(surveys, species_id != "DS")
```

We can also filter by multiple conditions:

```{r}
# temp_table <- filter(surveys, species_id == "DS", year >= "1998")
# tail(temp_table, 100)
```

The logical operator '&' and '|' mean "AND" and "OR"

```{r}
# filter(surveys, species_id == "DS" & year >= "1998")
```

```{r}
# filter(surveys, species_id == "DS" | year >= "1998")
```

Let's filter for values that are NA, we need the function 'is.na'

```{r}
# is.na(surveys)
# surveys$weight
# filter(surveys, !is.na(weight))
```

# Load surveys.csv into R using read.csv().

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
```

# Use select() to create a new data frame object with just the year, month, day, and species_id columns in that order.

```{r}
surveys_new <- select(surveys, year, month, day, species_id)
```

# Use mutate(), select(), and filter() with !is.na() to create a new data frame with the year, species_id, and weight in kilograms of each individual, with no null weights. The weight in the table is given in grams so you will need to create a new column for weight in kilograms by dividing the weight column by 1000.

```{r}
surveys_subset <- filter(surveys, !is.na(weight))
surveys_subset <- mutate(surveys_subset, weight_kg = weight/1000)
surveys_subset <- select(surveys_subset, year, species_id, weight_kg)
filter(surveys_subset, species_id == "SH")
```
# Pipes

`%>%` - `magrittr`  package required
'|>'  - inbuilt in R, but very new,s o you need an updated version of RStudio

```{r}
mean(x = c(1, 3, 65, 38, 100)) ## sequential code
c(1, 3, 65, 38, 100) %>% mean()

sqrt(mean (x = c(1, 3, 65, 38, 100)))

c(1, 3, 65, 38, 100) %>% mean () %>% sqrt()
```
## Use mutate(), select(), and filter() with is.na() to create a new data frame with the year, species_id, and weight in kilograms of each individual, with no null weights.

```{r}
surveys1 <- surveys %>% 
  mutate(weight_kg = weight/1000) %>%
  select(year, species_id, weight_kg) %>%
  filter(!is.na(weight_kg))
colnames(surveys1)
is.na(surveys1$weight_kg) %>%
  any()
```

## Use the filter() and select() to get the year, month, day, and species_id columns for all of the rows in the data frame where species_id is SH.

```{r}
surveys2 <- surveys %>% 
  select(year, month, day, species_id, weight) %>%
  filter(species_id == "SH") %>%
  filter(!is.na(weight)) %>%
  select(year, month, day, species_id)

surveys2$species_id == "SH"
surveys[surveys2$specis_id == "SH",]
colnames(surveys2)
nrow(surveys2)
```

## Obtain the data for only DS, sorted by year, with only the year and and weight columns

```{r}
ds_weight_by_year <- surveys %>%
  filter(species_id == "DS") %>%
  filter(!is.na(weight)) %>%
  arrange(year) %>%
  select(year, weight)
nrow(ds_weight_by_year)
```

## Data Aggregation

```{r}
surveys_by_year <- group_by(surveys, year)

surveys_by_year
```
```{r}
summarize(surveys_by_year, abundance = n())
```


## Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID.

```{r}
surveys_by_species_id <- surveys %>%
  group_by(species_id) %>%
  summarize(abundance = n())
surveys_by_species_id
```

## Use the group_by() and summarize() functions to get a count of the number of individuals in each species ID in each year.
```{r}
surveys_by_species_id_and_year <- surveys %>%
  group_by(species_id, year) %>%
  summarize(abundance = n())
surveys_by_species_id_and_year
```

## Use the filter(), group_by(), and summarize() functions to get the mean mass of species DO in each year.
```{r}
do_mean_mass_by_year <- surveys %>%
  filter(species_id == "DO") %>%
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  summarize(mean_weight_by_year = mean(weight)) 
do_mean_mass_by_year
```
# Setup your RStudio project

First, load the data, then look at it:

```{r}
surveys <- read.csv("../data-raw/surveys.csv")
species <- read.csv("../data-raw/species.csv")
plots <- read.csv("../data-raw/plots.csv")
View(surveys)
View(species)
View(plots)
```

```{r}
inner_join(surveys, species, by = "species_id")
left_join(surveys, species, by = "species_id")
right_join(surveys, species, by = "species_id")
full_join(surveys, species, by = "species_id")
```

How do we find out if we have matching columns in two tables

```{r}
surveys_colnames <- colnames(surveys)
species_colnames <- colnames(species)
surveys_colnames[7]
to_keep <- c(1,4,3,2)
to_keep <- c(FALSE,TRUE,TRUE,FALSE)
species_colnames[to_keep]
```

The %in% operator 

```{r}
1 %in% c(1,2,3)
c(1,2,3) %in% 1
colnames(species) %in% colnames(surveys)
matching_columns <- colnames(species) %in% colnames(surveys)
matching_columns
colnames(species)[matching_columns]
```

How to join more than 2 tables?

```{r}
combined <- inner_join(surveys, species, by = "species_id")
combined_full <- inner_join(combined, plots, by = "plot_id")
head(combined)
head(combined_full)
```

Run the code above using a pipe?

```{r}
combined_full <- surveys %>%
  inner_join(species, by = "species_id") %>%
  inner_join(plots, by = "plot_id")
head(combined_full)
```

## Use inner_join() and filter() to get a data frame with the information from the surveys and plots tables where the "plot_type" is "Control".

```{r}
control_plot_survey <- surveys %>%
  inner_join(plots, by = "plot_id") %>%
  filter(plot_type == "Control")
control_plot_survey
```

# We want to do an analysis comparing the size of individuals on the "Control" plots to the "Long-term Krat Exclosures".

## Create a data frame with the "year", "genus", "species", "weight" and "plot_type" for all cases where the
## plot type is either "Control" or "Long-term Krat Exclosure".
## Only include cases where "Taxa" is "Rodent".
## Remove any records where the “weight” is missing.

```{r}
colnames(species)
combined_full <- inner_join(surveys, species, by = "species_id") %>%
inner_join(plots, by = "plot_id") 
colnames(combined_full)
combined_full %>% 
select(year, genus, species, weight, plot_type, taxa) %>%
filter(plot_type == "Control" | plot_type == "Long-term Krat Exclosure") %>%
  filter(taxa == "Rodent") %>% 
  select(year, genus, species, weight, plot_type) 
```

# Solo In-Class Exercise

## Create a data frame with only data for the “species_id” “DO”, with the columns "year", "month", "day", "species_id", and "weight".

```{r}
combined_full %>%
  filter(species_id == "DO") %>%
  select(year, month, day, species_id, weight) -> data1
```

## Create a data frame with only data for species IDs "PP" and "PB" and for years starting in 1995, with the columns "year", "species_id", and "hindfoot_length", with no missing values for "hindfoot_length".

```{r}
combined_full %>% 
  filter(species_id == "PP" | species_id == "PB") %>%
  filter(year >= 1995) %>% 
  filter(!is.na(hindfoot_length)) -> data2
```

## Create a data frame with the average "hindfoot_length" for each "species_id" in each "year" with no null values.

```{r}
combined_full %>% 
  filter(!is.na(hindfoot_length)) %>%
  group_by(year, species_id) %>%
  summarize(mean_hindfoot_length_by_year = mean(hindfoot_length)) -> data3
```

## Create a data frame with the "year", "genus", "species", "weight" and "plot_type" for all cases where the "genus" is "Dipodomys".

```{r}
combined_full %>% 
  filter(genus == "Dipodomys") %>% 
  select(year, genus, species, weight, plot_type) -> data4
```

## Make a scatter plot with "weight" on the x-axis and "hindfoot_length" on the y-axis. Use a log10() scale on the x-axis. Color the points by "species_id". Include good axis labels.

```{r}
library("ggplot2")
ggplot(data = combined_full, mapping = aes(x = weight, y = hindfoot_length, color = species_id)) +  
  geom_point() +
  scale_x_log10() +
  labs(x = "Weight", y = "Hindfoot Length", title = "Data from Surveys, Plots, and Species")
```

## Make a histogram of weights with a separate subplot for each "species_id". Do not include species with no weights. Set the "scales" argument to "free_y" so that the y-axes can vary. Include good axis labels.

```{r}
combined_full %>% 
  filter(!is.na(weight)) %>%
  ggplot(data = combined_full, mapping = aes(x = weight)) + 
   labs(x = "Weight", y = "Number of Individuals in the Species", title = "Distribution of Weight per Species") + 
  geom_histogram() +
  facet_wrap(~species_id, scales = "free_y")
```

## Make a plot with histograms of the weights of three species, "PP", "PB", and "DM", colored by "species_id", with a different facet (i.e., subplot) for each of three "plot_type"’s "Control", "Long-term Krat Exclosure", and "Short-term Krat Exclosure". Include good axis labels and a title for the plot. Export the plot to a PNG file.

```{r}
combined_full %>% 
  filter(!is.na(weight)) %>% 
  filter(species_id == "PP" | species_id == "PB" | species_id == "DM") %>% 
  filter(plot_type != "Rodent Exclosure" & plot_type != "Spectab exclosure") %>% 
ggplot(mapping = aes(x = weight, color = species_id, fill = species_id)) +
  geom_histogram() +
   labs(x = "Weight", y = "Number of Individuals in the Species", title = "Distribution of Weight per Species by Plot Type")
  facet_wrap(~plot_type)
```

# Solo In-Class Exercise

## Develop a data manipulation pipeline for the Portal surveys table that produces a table of data for only the three Dipodomys species ("DM", "DO", "DS").

## The species IDs should be presented as lower case, ## not upper case.
## The table should contain information on the date, the species ID, the weight and hindfoot length.
## The data should not include null values for either weight or hindfoot length.
## The table should be sorted first by the species (so that each species is grouped together) and then by weight, with the largest weights at the top.

```{r}
combined_full %>% 
  mutate(species_id = tolower(species_id)) %>% 
  select(month, day, year, species_id, weight, hindfoot_length) %>% 
  filter(species_id == "dm" | species_id == "do" | species_id == "ds") %>% 
  filter(!is.na(weight)) %>% 
  filter(!is.na(hindfoot_length)) %>% 
  arrange(species_id) %>% 
  arrange(desc(weight))
```




