---
title: "uhuru dataset"
author: "Jai Jot Kaur"
date: "2022-10-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      root.dir =
"Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw")
```

### 1. Describing the data that we are using

This experiment characterizes the effects of selectively removing the class of herbivores of the largest size from the system and determining how the direction and  the magnitude of the effects take shape from the variation that is seen in the precipitation regimes. There were three herbivore-exclusion treatments that were electrically fences and there was an unfensed control which were applied to contiguous 1-ha plot blocks. The fences treatments included "Mega", "Meso", and "Total." Mega had the exclusion of just the elephants and giraffes, Meso had the exclusion of both the megaherbivores and the mesoherbivores which were 40 kilograms or larger, and Total excludes all of the herbivores that were equal to or over 5 kilograms.  Replication of each of the blocks occurred three times at three various points of the 20 kilometer rainfall gradient and the data was presented spanning the years 2008 and 2013. 

We are using the dataset from [this study](https://esapubs.org/archive/ecol/E095/064/metadata.php)

Add a picture of an acacia

### 2. Reading the data table into R

First make sure that we are in the correct working directory, we use function 'getwd()'.

[1] "/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw"
in the setup chunk above.

```{r}
acacia <- read.csv(file = "/Users/jaijotkaur/Desktop/BIO197/data_science_research/data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t")
```

### 3. Explore our data set

```{r}
acacia
head(acacia)
summary(acacia)
colnames(acacia)
rownames(acacia)
class(acacia$SURVEY)
# the sapply function allows to apply a function to a list of objects
# a data frame is a list of vectors of the same length
sapply(acacia, class)
```

How do I test where the computer is?

```{r}
getwd()
r_proj_wd <- "/Users/jaijotkaur/Desktop/BIO197/data_science_research"
r_chunk_wd <- getwd()
r_proj_wd == r_chunk_wd
r_chunk_wd
```

Make sure that everything that is a number is actually numeric.

One way to do this is with the function 'summary', and checking the type of data at each column visually

Another way is using the type function

```{r}
typeof(acacia[,"HEIGHT"])

acacia$HEIGHT
```

We identified a colimn that has problematic data. We need to fix it!

We are going to read the data table again, but we are gonna assign 'NA' to the "dead" value that we do not want in our "HEIGHT" column



## Cleaning our raw data
### 3.2 Assigning "NA" to missing/incorrect values

```{r}
acacia <- read.csv(file = "../data-raw/ACACIA_DREPANOLOBIUM_SURVEY.txt", sep = "\t", na.strings = "dead")
```

Let's check if this worked!

```{r}
typeof(acacia$HEIGHT)
class(acacia$HEIGHT)
```

### 4. Visualize our data

For this, we are using the 'ggplot' package.
Let's install it and load it:

```{r}
# install.packages("ggplot2")
library("ggplot2")
```

Now we are gonna create our first plotting layer with the function 'ggplot'.

```{r}
colnames(acacia)
acacia$CIRC
```

The ggplot function creates a blank canvas, and the canvas contains our data and the variables that we want to plot.

A scatter plot can be created with the function `geom_point()`

```{r}
colors()
```

```{r}
?pch
```

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 1, color = "peachpuff", alpha = 0.8, shape = 11) +
  labs(x = "Circumference (cm)", y = "Height (cm)", title = "Data from UHURU Acacia Survey")
```

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 1, color = "blue", alpha = 0.1) +
  labs(x = "Circumference (cm)", y = "Height (cm)", title = "Data from UHURU Acacia Survey")
```
To rescale the plotting of the axis to log scale we use the function 'scale_y_log_10()'

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) +
  geom_point(size = 1, color = "purple", alpha = 0.4) + labs(x = "Circumference (cm)", y = "Height (cm)", title = "Data from UHURU Acacia Survey") +
  scale_x_log10() + scale_y_log10()
```

We have the information on experimental treatment in the treatment column
```{r}
acacia$TREATMENT
getwd()
```

Let's add information on treatment to our plot:

```{r}
ggplot(data = acacia, mapping = aes (x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point(size = 2, alpha = 0.5)
```

# Subplots or facets

The function to creat subplots by a third variable is called 'facet_wrap'
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT, color = TREATMENT)) + 
         geom_point() +
         facet_wrap(~TREATMENT)
```
# A little interpretation of what is going on here
This graph shows the effects of the treatment in the areas that were restructed to certain animals. The area that had been completely open with no fencing, and, therefore used as a control, showed very little trees with only a point demonstrating one that was very thin and short. The one that blocked off all herbivores was the total; this showed there to be many trees. For mega there were trees that were shorter, while meso had trees that were taller. This makes sense because when the taller animals are blocked out, they cannot consume the leaves at the top, while when the smaller animals are blocked out, the trees remain tall. When there is complete fencing for most of the herbivores, there is many trees due to their being no consumption. The meso and mega demonstrate a slight linear pattern in the start and the kind of branch off to points that are more random.

## Model fitting functions

The `geom_smooth()' fucntion allows to fit linear models to a set of points

```{r}
ggplot(data = acacia, mapping = aes(x= CIRC, y = HEIGHT, color = TREATMENT, linetype = TREATMENT)) + geom_point() + 
  geom_smooth(method = "loess") + 
  facet_wrap(~TREATMENT)
```

## Histograms and barplots

For bar plots we use the geom_bar() function:

```{r}
ggplot(data = acacia, aes(x = TREATMENT)) +
  geom_bar()
```

If we want to see the distribution of a continuous variable we use the geom_histogram() function:

```{r}
ggplot(data = acacia, mapping = aes(x = HEIGHT, fill = TREATMENT)) + 
  geom_histogram(bins = 20, color = "black")
```

## Layer multiple data from the same or different data sets

```{r}
ggplot() +
  geom_point(data = acacia, 
             mapping = aes(x = CIRC, y = HEIGHT,
                           color = TREATMENT)) +
  geom_smooth(data = acacia,
              mapping = aes(x =CIRC, y = HEIGHT)) +
  geom_histogram(data = acacia,
                 mapping = aes(x = CIRC, color = TREATMENT), alpha = 0.2)

```

# Save images as files

```{r}
ggsave(filename = "acacia_by_treatment.pdf", width = 5, height = 3)
```

# Exercise 1. Scatterplots

## Make a scatter plot with CIRC on the x axis and AXIS1 (the maximum canopy width) on the y axis. Label the x axis “Circumference” and the y axis “Canopy Diameter”.
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1)) +
  geom_point() +
  labs(x = "Circumference", y = "Canopy Diameter", title = "Data from UHURU Acacia Survey")
```

## The same plot as (1), but with both axes scaled logarithmically (using scale_x_log10 and scale_y_log10).
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1)) +
  geom_point() +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Circumference", y = "Canopy Diameter", title = "Data from UHURU Acacia Survey")
```

## The same plot as (1), but with points colored based on the ANT column (the species of ant symbiont living with the acacia)
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1, color = ANT)) +
  geom_point() +
  labs(x = "Circumference", y = "Canopy Diameter", title = "Data from UHURU Acacia Survey")
```

## The same plot as (3)), but instead of different colors show different species of ant (values of ANT) each in a separate subplot.
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1, color = ANT)) +
  geom_point() +
  labs(x = "Circumference", y = "Canopy Diameter", title = "Data from UHURU Acacia Survey") + 
  facet_wrap(~ANT)
```

## The same plot as (4) but add a simple model of the data by adding geom_smooth.
```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = AXIS1, color = ANT)) +
  geom_point() +
  labs(x = "Circumference", y = "Canopy Diameter", title = "Data from UHURU Acacia Survey") + 
  geom_smooth() +
  facet_wrap(~ANT)
```

# Exercise 2. Histograms

## Make a bar plot of the number of acacia with each mutualist ant species (using the ANT column).
```{r}
ggplot(data = acacia, aes(x = ANT)) +
  geom_bar()
```

## Make a histogram of the height of acacia (using the HEIGHT column). Label the x axis “Height (m)” and the y axis “Number of Acacia”.
```{r}
ggplot(data = acacia, mapping = aes(x = HEIGHT)) + 
   labs(x = "Height (m)", y = "Number of Acacia", title = "Data from UHURU Acacia Survey") + 
  geom_histogram()
```

## Make a plot that shows histograms of both AXIS1 and AXIS2. Due to the way the data is structured you’ll need to add a 2nd geom_histogram() layer that specifies a new aesthetic. To make it possible to see both sets of bars you’ll need to make them transparent with the optional argument alpha = 0.3. Set the color for AXIS1 to “red” and AXIS2 to “black” using the fill argument. Label the x axis “Canopy Diameter(m)” and the y axis “Number of Acacia”.
```{r}
ggplot()+
  geom_histogram(data = acacia, mapping = aes(x = AXIS1), alpha = 0.3, color = "red") + 
  geom_histogram(data = acacia, mapping = aes(x = AXIS2), alpha = 0.3, color = "black") +
   labs(x = "Canopy Diameter (m)", y = "Number of Acacia", title = "Data from UHURU Acacia Survey")
```

## Use facet_wrap() to make the same plot as (3) but with one subplot for each treatment. Set the number of bins in the histogram to 10.
```{r}
ggplot()+
  geom_histogram(data = acacia, mapping = aes(x = AXIS1), alpha = 0.3, color = "red", bins = 10) + 
  geom_histogram(data = acacia, mapping = aes(x = AXIS2), alpha = 0.3, color = "black", bins = 10) +
   labs(x = "Canopy Diameter (m)", y = "Number of Acacia", title = "Data from UHURU Acacia Survey") +
  facet_wrap(~~TREATMENT) 
```

