---
title: "vectors and data frames"
author: "Jai Jot Kaur"
date: "10/20/2022"
output: html_document
---

# The relationship between vectors and data frames

## How to create vectors

```{r}
a_number <- 100
length(a_number)
c(1,2,3)
c(FALSE, TRUE, TRUE)
c("Luna", "Jai", "Azul", "Marcos")
1:4
0.5:5
c(1, 6, 7, 100, -1)
1;100
long_vector <- 1:546378
length(long_vector)
tail(long_vector)
seq(1, 100, 0.3)
seq(-100, 5, 2)
100:20
-100:-5
rep(1:5, 3)
rep("luna", 100)
rep(c(TRUE, FALSE), 7)
letters
LETTERS
```


## Joining vectors to create data frames

```{r}
year_data <- data.frame(names = c("Azul", "Jai", "Marcos"), 
                        year = c(4,3,4),
                        college = "UC Merced")
year_data
```

# Joint in-class exercise

```{r}
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)

taxus_baccata_data <- data.frame(length = length,
                                 width = width,
                                 height = height,
                                 genus = "Taxus",
                                 species = "baccata")
taxus_baccata_data
```

## Extracting values from vectors and data frames

```{r}
surveys <- read.csv(file = "../data-raw/surveys.csv")
head(surveys)
# extracting one column at a time, we use the dollar sign 
surveys$year
surveys$month
surveys$species_id
# if we want to extract rows and/or multiple columns, we use the square brackets
surveys[1,2]
surveys[1:5,]
```

# Exercise

## Use $ to extract the weight column into a vector called surveys_weight
```{r}
surveys_weight <- surveys$weight
surveys_weight
```

## Use [] to extract the month column into a vector called surveys_month
```{r}
surveys_month <- surveys[, "month"]
surveys_month
```

## Extract the hindfoot_length column into a vector and calculate the mean hindfoot length ignoring missing values.
```{r}
surveys_filtered_hl <- filter(surveys, !is.na(hindfoot_length))
surveys_hindfoot_length <- surveys_filtered_hl[, "hindfoot_length"]
mean(surveys_hindfoot_length)

surveys_hindfoot <- surveys$hindfoot_length
mean(surveys_hindfoot, na.rm = TRUE)
```

### Extraxt values from vectors

```{r}
letters[1:5]
letters[c(17, 12, 9, 3)]
letters[5:1]
letters[-c(1, 6, 8, 9)]
colnames(surveys)
surveys[, -(1:4)]
```




